Spin-orbit-coupling-induced squeezing in the optical lattice clock

### 2018-07-09 soc_squeezing/soc_squeezing_notes.tex
removed mean-field treatment changed figure directory from 'plots' to 'figures' added OAT and TAT comparison figure

%%% figures
\usepackage{graphicx} % for figures
\usepackage{grffile} % help latex properly identify figure extensions
\usepackage[caption=false]{subfig} % subfigures (via \subfloat[]{})
\graphicspath{{./plots/}} % set path for all figures

@@

\section{Mean-field treatment}

To sum up our developments thus far, spin-orbit coupling gives rise to
an effective inhomogeneous field, while $s$-wave interactions give
rise to a collective spin interaction, which means that the system is
characterized by the Hamiltonian \note{[MP: I removed the mention of
  non-uniform coupling (i.e.~through $\xi_{ij}$), as I feel it is
  unnecessary given the preceding section on the collective spin
  model.  The uniform all-to-all coupling in \eqref{eq:H_int} is exact
  under the assumptions of the spin model, so there is no need to
  ``re-derive'' it by throwing out inhomogeneous couplings and taking
  $\xi_{ij}\to\xi\equiv\overline{\xi_{ij}}$.]}
\begin{align}
  H = H_{\t{int}} + H_{\t{SOC}} = -\xi\v S\c\v S - \sum_q h_q s_q^z,
  \label{eq:H_full}
\end{align}

\note{[MP] Note: ``eqnarray'' environment is generally not recommended
  for typesetting math, as it has inconsistent spacing in addition to
  some other problems (discussed a bit
  \href{http://www.tug.org/pracjourn/2006-4/madsen/madsen.pdf}{here},
  if you're interested).  The ``align'' environment, on the other
  hand, is provided by math packages for the purposes of typesetting
  math, lacks spacing problems, and generally has more functionality.}

This collective Heisenberg model is a celebrated model for magnetism,
and is closely related to the Richardson-Gaudin models for
superconductivity based on the Anderson pseudo-spin mapping
\note{[cite?]}.  So before demonstrating how to generate an effective
one-axis squeezing Hamiltonian based on Eqn.~\eqref{eq:H_full}, we
first try to develop better understanding of this Hamiltonian.  In
fact, this model is well-studied at the mean-field (MF) level as given
by
\begin{align}
  H_{\t{MF}} = -\sum_q \v s_q \c \v B_q,
  &&
  \v B_q \equiv \xi\v{\mathcal S} + \v h_q,
  &&
  \v{\mathcal S} \equiv \bk{\v S},
  &&
  \v h_q \equiv \p{0,0,h_q}.
  \label{eq:mean_field}
\end{align}
\note{[MP: I removed a factor of 2 in front of $\xi$, as I believe it
  was a mistake.  Was this correct?  If the factor of 2 was indeed
  incorrect, please make sure that there are no extra factors of 2 in
  Peiru's codes.]}
% MP: as an additional side note, I personally recommend the use of
% \braket{X} or (\bk{X}, which I defined as a shorthand) in place of
% \langle X \rangle; I find this is a little easier to read.
% Similarly, you can use \ket{X} (\bra{X}) in place of |X\rangle in
% place of (\langle X|), and \op{X}{Y} in place of |X\rangle\langle Y|
% (where "op" stands for "outer product")

The behavior of this model is determined by a competition between the
non-local MF coupling through $\xi$ and an effective inhomogeneous
axial field $\v h_q$ whose strength has a variance
$8J_0^2\sin^2\p{\phi/2}$.  Fig.~\ref{fig:MFcontrast} shows the
dynamics of the system as computed by numerical simulation of
\eqref{eq:mean_field}, after initialization with all spins pointing
along the $x$ axis, i.e.~in the state
$\bigotimes_q\p{\ket{q,\g}+\ket{q,\e}}/\sqrt2$. \note{[MP: I added
  text specifying the initial state; is it correct?]}  As the
interaction strength $\xi$ increases, the contrast
$\mathcal C\equiv2\abs{\mathcal S_+}/N$, \note{[MP: I propose writing
  $\mathcal C\equiv\sqrt{\mathcal S_x^2+\mathcal S_y^2}/\p{N/2}$;
  while longer, I feel like it makes the physical interpretation of
  $\mathcal C$ more clear.]} i.e.~the normalized magnitude of the
collective spin vector in the $x$-$y$ plane, gets stabilized against
dephasing induced by the effective inhomogeneous axial field $\v h_q$.
Furthermore, when $\xi$ is below a critical value $\xi_{\t{c}}$, the
contrast exhibits some oscillatory behavior, but ultimately decays at
long times.  Above the critical value $\xi_{\t{c}}$, however, the
contrast reaches a non-zero steady state, with oscillatory behavior
that is damped with increasing $\xi$.  These two phases can be
understood through the underlying integrability of
\eqref{eq:mean_field}.  Using the techniques in \note{[cite]}, the
steady state contrast $\bar{\mathcal C}$ at the MF level can be
obtained analytically as
\begin{align}
  \bar{\mathcal C} =
  \begin{cases}
    \sqrt{1-\sp{4J_0\sin\p{\phi/2}/N\xi}^2},
    & \text{if}~ \abs{4J_0\sin(\phi/2)/N\xi} < 1 \\
    0,
    & \text{otherwise}
  \end{cases}
\end{align}

\begin{figure}
\centering
\includegraphics[width=3.5in]{MFcontrast.pdf}
\caption{
The contrast as a function of time and interaction strength.
}
\label{fig:MFcontrast}
\end{figure}

Comparing the MF simulation results with the quantum simulation
results \note{[MP: In my experience, ``quantum simulation'' refers to
  that which is performed with a quantum simulator, i.e.~via
  experiment.  I assume that's not what you meant here, and that these
  are actually some sort of Monte Carlo / quantum trajectory / TWA
  simulations.  Can we specify the method?]}, however, we find that
the contrast decreases rather than remain as a constant even if
$\xi\gg\xi_{\t{c}}$, as shown in Fig.~\ref{fig:EffectiveSqueezing}.

\begin{figure}
\centering
\includegraphics[width=3in]{Fig33.pdf}
\caption{Contrast comparison among the MF simulation results, quantum simulation results and analytic expressions:
(a) Good agreement between MF simulation (solid line) and quantum simulation (dashed line) at short time,
while  the discrepancy occurs at longer time due to the quantum fluctuations.
(b-c) Benchmark the approximated one-axis twisting analytical expressions (dashed line) with the quantum treatment (solid line).
}
\label{fig:EffectiveSqueezing}
\end{figure}

\note{[MP: While it may be nice for understanding some properties of
  the Hamiltonian in \eqref{eq:H_full}, I do not think that this text
  about contrast and mean-field treatment adds much to our goal of
  spin squeezing.  Do we need to include this material here?  I would
  cut this section out for now.]}

@@

Rather than pulsing the clock laser, we can apply a continuous drive
to realize the Hamiltonian
\begin{align}
  H_{\t{drive}} = H_{\t{free}} + H_{\t{clock}}
  = -\xi\v S\c\v S - \bar h S_z - \Omega S_x - \sum_n b_n s_n^z,
\end{align}
where the drive strength $\Omega$ may generally depend on time.  As
shown in Appendix \ref{sec:effective_squeezing_drive}, the addition of
a drive with $\abs{\Omega}\ll N\abs{\xi}$ does not modify the
squeezing behavior of the spin-orbit coupling terms in
$H_{\t{drive}}$, whose effective action on the $S=N/2$ manifold is
captured by
\begin{align}
  H_{\t{drive,eff}} = -\bar h S_z - \Omega S_x + \chi S_z^2.
  \label{eq:H_drive_eff}
\end{align}
To follow the continuous-drive two-axis twisting prescription in
ref.~\cite{huang2015twoaxis}, we now
\begin{enumerate*}[label=(\roman*)]
\item blue-detune the clock laser by $\bar h$ and shift into an
  appropriate rotating frame to eliminate the $\sim S_z$ term in
  \eqref{eq:H_drive_eff},
\item modulate the drive as $\Omega\p{t}=\beta\omega\cos\p{\omega t}$
  with $\omega\gg N\chi$, and
\item move into the rotating frame of $-\Omega\p{t}S_x$.
  % , and finally
% \item shift $H_{\t{drive,eff}}$ by the constant
  % $\p{\chi/2}\sp{\J_0\p{2\beta}-1} S^2$, where $\J_0\p{x}$ is the
  % zero-order Bessel function of the first kind evaluated at $x$.
\end{enumerate*}
After a secular approximation which relies on $\omega\gg N\chi$, this
procedure results in the effective Hamiltonian
\begin{align}
  H_{\t{drive,eff}}^{\t{mod}}
  = \f{\chi}{2} \p{\sp{\J_0\p{2\beta}+1} S_z^2
    - \sp{\J_0\p{2\beta}-1} S_y^2},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
Choosing a modulation index $\beta$ such that $\J_0\p{2\beta}=1/3$ or
$\J_0\p{2\beta}=-1/3$ then respectively results in the two-axis
twisting Hamiltonians
\begin{align}
  H_{\t{TAT}}^{z,x}
  = \f{\chi}{3} \p{2 S_z^2 + S_y^2}
  \simeq \f{\chi}{3} \p{S_z^2 - S_x^2},
  &&
  H_{\t{TAT}}^{y,x}
  = \f{\chi}{3} \p{S_z^2 + 2 S_y^2}
  \simeq \f{\chi}{3}\p{S_y^2 - S_x^2},
  \label{eq:H_TAT_drive}
\end{align}
where $\simeq$ denotes equality up to a global energy shift. Given the
set of parameters in Table \ref{tab:parameters}, the two-axis twisting
Hamiltonian $H_{\t{TAT}}^{z,x}$ can be achieved with
$\omega=20\times2\pi$ Hz and $\beta=0.906$.  The optimal squeezing
time for this protocol is $t_{\t{opt}}^{\t{TAT}}\approx0.98$ seconds.

>>>>

%%% figures
\usepackage{graphicx} % for figures
\usepackage{grffile} % help latex properly identify figure extensions
\usepackage[caption=false]{subfig} % subfigures (via \subfloat[]{})
\graphicspath{{./figures/}} % set path for all figures

@@

@@

Rather than pulsing the clock laser, we can apply a continuous drive
to realize the Hamiltonian
\begin{align}
  H_{\t{drive}} = H_{\t{free}} + H_{\t{clock}}
  = -\xi\v S\c\v S - \bar h S_z - \Omega S_x - \sum_n b_n s_n^z,
\end{align}
where the drive strength $\Omega$ may generally depend on time.  As
shown in Appendix \ref{sec:effective_squeezing_drive}, the addition of
a drive with $\abs{\Omega}\ll N\abs{\xi}$ does not modify the
squeezing behavior of the spin-orbit coupling terms in
$H_{\t{drive}}$, whose effective action on the $S=N/2$ manifold is
captured by
\begin{align}
  H_{\t{drive,eff}} = -\bar h S_z - \Omega S_x + \chi S_z^2.
  \label{eq:H_drive_eff}
\end{align}
To follow the continuous-drive two-axis twisting prescription in
ref.~\cite{huang2015twoaxis}, we now
\begin{enumerate*}[label=(\roman*)]
\item blue-detune the clock laser by $\bar h$ and shift into an
  appropriate rotating frame to eliminate the $\sim S_z$ term in
  \eqref{eq:H_drive_eff},
\item modulate the drive as $\Omega\p{t}=\beta\omega\cos\p{\omega t}$
  with $\omega\gg N\chi$, and
\item move into the rotating frame of $-\Omega\p{t}S_x$.
\end{enumerate*}
After a secular approximation which relies on $\omega\gg N\chi$, this
procedure results in the effective Hamiltonian
\begin{align}
  H_{\t{drive,eff}}^{\t{mod}}
  = \f{\chi}{2} \p{\sp{\J_0\p{2\beta}+1} S_z^2
    - \sp{\J_0\p{2\beta}-1} S_y^2},
\end{align}
where $\J_0$ is the zero-order Bessel function of the first kind.
Choosing a modulation index $\beta$ such that $\J_0\p{2\beta}=1/3$ or
$\J_0\p{2\beta}=-1/3$ then respectively results in the two-axis
twisting Hamiltonians
\begin{align}
  H_{\t{TAT}}^{z,x}
  = \f{\chi}{3} \p{2 S_z^2 + S_y^2}
  \simeq \f{\chi}{3} \p{S_z^2 - S_x^2},
  &&
  H_{\t{TAT}}^{y,x}
  = \f{\chi}{3} \p{S_z^2 + 2 S_y^2}
  \simeq \f{\chi}{3}\p{S_y^2 - S_x^2},
  \label{eq:H_TAT_drive}
\end{align}
where $\simeq$ denotes equality up to a global energy shift. Given the
set of parameters in Table \ref{tab:parameters}, the two-axis twisting
Hamiltonian $H_{\t{TAT}}^{z,x}$ can be achieved with
$\omega=20\times2\pi$ Hz and $\beta=0.906$.  The optimal squeezing
time for this protocol is $t_{\t{opt}}^{\t{TAT}}\approx0.98$ seconds.
A comparison of squeezing through $H_{\t{OAT}}$ and
$H_{\t{TAT}}^{z,x}$ via the parameters in Table \ref{tab:parameters}
is provided in Figure \ref{fig:squeezing_comparison}.

\begin{figure}[h]
  \centering
  \includegraphics{squeezing_comparison.pdf}
  \caption{Comparison of squeezing through $H_{\t{OAT}}$ and
    $H_{\t{TAT}}^{z,x}$ via the parameters in Table
    \ref{tab:parameters} and the continuous two-axis twisting protocol
    with $\omega=20\times2\pi$ Hz and $\beta=0.906$.}
  \label{fig:parameters}
\end{figure}
